---
title: "Untitled"
output:
  word_document: default
  html_document: default
---


```{r}
statistics <- read.csv("/Users/caseysun/Downloads/Full_Dataset.csv")
head(statistics)
```



```{r}
# Install and load the countrycode package
if (!requireNamespace("countrycode", quietly = TRUE)) {
  install.packages("countrycode")
}

# Load the countrycode package
library(countrycode)

# Assuming your data frame is named 'statistics', and the country column is named 'Country'
statistics$Continent <- countrycode(sourcevar = statistics$Country,
                                    origin = "country.name",
                                    destination = "continent")


# Place Continent after Country colum
statistics <- statistics[, c("Country", "Continent", names(statistics)[-c(which(names(statistics) %in% c("Country", "Continent")))])]

# Display the updated data frame
head(statistics)
```

```{r}
#The function grouped all North and South American countries as Americas 
# Also, the function could not identify the Continent for Micronesia, therefore we manually hard-coded these cases.

# Manually override the continent for specific cases
statistics$Continent[statistics$Country == "Micronesia"] <- "Oceania"
statistics$Continent[statistics$Country %in% c("United States", "Canada", "Mexico")] <- "North America"
statistics$Continent[statistics$Country %in% c("Brazil", "Argentina", "Colombia", "Chile", "Peru", "Ecuador", "Bolivia", "Paraguay", "Trinidad                                                 and Tobago", "Uruguay", "Venezuela", "Guyana", "Suriname", "French Guiana")] <- "South America"
statistics$Continent[statistics$Country %in% c("Guatemala", "Panama", "Costa Rica", "El Salvador", "Honduras", "Nicaragua", "Belize")] <-                                                      "Central America"
statistics$Continent[statistics$Country %in% c("Dominica", "Saint Kitts & Nevis", "Jamaica", "St. Vincent & Grenadines", "Trinidad and Tobago", 
                                               "Haiti", "Bahamas","Dominican Republic", "Barbados", "Saint Lucia", "Antigua and Barbuda",                                                      "Grenada")] <- "Caribbean"
```


```{r}
#View(statistics)
```



```{r}
str(statistics)
```



```{r}
# Custom function to convert Export and Import values with trillion, billion, million to numeric
convert_to_numeric <- function(value) {
  # Extract numeric part
  numeric_part <- as.numeric(gsub("[^0-9.]", "", value))

  # Extract unit part
  unit_part <- gsub("[0-9.]", "", value)

  # Convert to numeric based on unit
  if (grepl("trillion", unit_part, ignore.case = TRUE)) {
    return(numeric_part * 1e12)
  } else if (grepl("billion", unit_part, ignore.case = TRUE)) {
    return(numeric_part * 1e9)
  } else if (grepl("million", unit_part, ignore.case = TRUE)) {
    return(numeric_part * 1e6)
  } else {
    return(numeric_part)
  }
}

# Apply custom function to "Exports" and "Imports" columns
statistics$Exports <- sapply(statistics$Exports, convert_to_numeric)
statistics$Imports <- sapply(statistics$Imports, convert_to_numeric)

# Display the updated structure
str(statistics)
```



# Handling Missing Values

```{r}

# Load the necessary package
library(dplyr)

# Function to replace missing values with the median of the continent
replace_na_with_median <- function(df, col_name) {
  df %>%
    group_by({{ col_name }}) %>%
    mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
    ungroup()
}

# Replace missing values in numeric columns with the median of the continent
statistics <- replace_na_with_median(statistics, col_name = Continent)

statistics <- as.data.frame(statistics)

# Display the updated data frame
head(statistics)

```
```{r}
tail(statistics)
```

```{r}
# Set row names to "Country" column
row.names(statistics) <- statistics$Country

# Keep a copy of the dataset
mydata <- statistics

# Removing categorical columns
statistics <- statistics[, c(-1, -2, -20, -23, -25)] 

# Display the updated data frame
head(statistics)
```

```{r}
# reversing the values of some variables.
statistics$UnEmpRate = max(statistics$UnEmpRate) - statistics$UnEmpRate
statistics$IMR = max(statistics$IMR) - statistics$IMR
statistics$InfRate = max(statistics$InfRate) - statistics$InfRate
statistics$CO2 = max(statistics$CO2) - statistics$CO2
statistics$AQI = max(statistics$AQI) - statistics$AQI
head(statistics)
```



```{r}
#Scale data
s.statistics = scale(statistics)
```


```{r}
corr = cor(s.statistics)
round(corr, 2)
```
```{r}
unique_values <- unique(mydata$Continent)
unique_values
number_of_unique_values <- length(unique(mydata$Continent))
number_of_unique_values
```

```{r}
options(digits = 3)
```

## hierarchical clustering - single

```{r}
## hierarchical clustering - single
dist <- dist(s.statistics)
hc.single <- hclust(dist, 'single')
# cluster dendrogram
plot(hc.single, cex = 0.3, main = 'Cluster Dendrogram of HC.single')
```
```{r}
# find the number of clusters
plot(rev(hc.single$height), main = 'scree plot of HC.single')
```

Based on the scree plot, set h = 5  as the cut point.
```{r}
ct <- cutree(hc.single, h = 5)
table(ct)
```
This clustering outcome is not good.

```{r}
# cluster mean-HC.single
cluster1 = subset(rownames(s.statistics), ct == 1)
index1 = match(cluster1, rownames(s.statistics))
clust1mean = colMeans(s.statistics[index1, , drop = FALSE])

cluster2 = subset(rownames(s.statistics), ct == 2)
index2 = match(cluster2, rownames(s.statistics))
clust2mean = colMeans(s.statistics[index2, , drop = FALSE])

cluster3 = subset(rownames(s.statistics), ct == 3)
index3 = match(cluster3, rownames(s.statistics))
clust3mean = colMeans(s.statistics[index3, ])

cluster4 = subset(rownames(s.statistics), ct == 4)
index4 = match(cluster4, rownames(s.statistics))
clust4mean = colMeans(s.statistics[index4, , drop = FALSE])

cluster5 = subset(rownames(s.statistics), ct == 5)
index5 = match(cluster5, rownames(s.statistics))
clust5mean = colMeans(s.statistics[index5, ])

cluster6 = subset(rownames(s.statistics), ct == 6)
index6 = match(cluster6, rownames(s.statistics))
clust6mean = colMeans(s.statistics[index6, ])

cluster7 = subset(rownames(s.statistics), ct == 7)
index7 = match(cluster7, rownames(s.statistics))
clust7mean = colMeans(s.statistics[index7, , drop = FALSE])

cluster8 = subset(rownames(s.statistics), ct == 8)
index8 = match(cluster8, rownames(s.statistics))
clust8mean = colMeans(s.statistics[index8, , drop = FALSE])

clustMeans <- rbind(clust1mean, clust2mean, clust3mean, clust4mean, clust5mean, clust6mean, clust7mean, clust8mean)
library(knitr)
kable(clustMeans, caption = "Column Means for Each Cluster-HC.single")
```


```{r}
# Interpreting HC.single based on PCA
pca <- princomp(s.statistics)
pca$loadings[,1:2]

plot(pca$scores[, 1:2], col = 'white', main='PCA Biplot - HC.single')
text(pca$scores[, 1:2], labels=abbreviate(row.names(s.statistics)), col=ct, cex=0.5)
```
The plot is hard to interpret, hierarchical clustering single method is not working well.



```{r}
#install.packages('maps')
library(ggplot2)
library(maps)
world_map <- map_data("world")
country_clusters <- data.frame(country = names(ct), cluster = ct)
# match the country name
world_countries <- unique(world_map$region)
your_countries <- unique(country_clusters$country)

# Countries in your data not in the world map
not_in_world_map <- your_countries[!your_countries %in% world_countries]

```
```{r}
print(not_in_world_map)
```
```{r}
# make a map
country_clusters$country <- gsub("United States", "USA", country_clusters$country)
country_clusters$country <- gsub("United Kingdom", "UK", country_clusters$country)
country_clusters$country <- gsub("Czech Republic (Czechia)", "Czech Republic", country_clusters$country)
country_clusters$country <- gsub("Côte d'Ivoire", "Ivory Coast", country_clusters$country)
country_clusters$country <- gsub("DR Congo", "Democratic Republic of the Congo", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Trinidad", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Tobago", country_clusters$country)
country_clusters$country <- gsub("State of Palestine", "Palestine", country_clusters$country)
country_clusters$country <- gsub("Congo", "Republic of Congo", country_clusters$country)
country_clusters$country <- gsub("Eswatini", "Swaziland", country_clusters$country)
country_clusters$country <- gsub("Cabo Verde", "Cape Verde", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Antigua", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Barbuda", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Saint Kitts", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Nevis", country_clusters$country)
country_clusters$country <- gsub("St. Vincent & Grenadines", "Saint Vincent", country_clusters$country)
country_clusters$country <- gsub("Sao Tome & Principe", "Sao Tome and Principe", country_clusters$country)


world_map <- map_data("world")

world_map_clusters <- merge(world_map, country_clusters, by.x = "region", by.y = "country", all.x = TRUE)

ggplot(data = world_map_clusters) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(cluster))) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  theme_minimal() +
  labs(fill = "Cluster")
```

## hierarchical clustering - complete

```{r}
## hierarchical clustering - complete
hc.complete <- hclust(dist, 'complete')
# cluster dendrogram
plot(hc.complete, cex = 0.3, main = 'Cluster Dendrogram of HC.complete')
```
```{r}
# find the number of clusters
plot(rev(hc.complete$height), main='scree plot of HC.complete')

```
Based on the scree plot, set h = 8  as the cut point.
```{r}
ct2 <- cutree(hc.complete, h = 8)
table(ct2)
```
```{r}
# cluster mean-HC.complete
cluster1 = subset(rownames(s.statistics), ct2 == 1)
index1 = match(cluster1, rownames(s.statistics))
clust1mean = colMeans(s.statistics[index1, , drop = FALSE])

cluster2 = subset(rownames(s.statistics), ct2 == 2)
index2 = match(cluster2, rownames(s.statistics))
clust2mean = colMeans(s.statistics[index2, , drop = FALSE])

cluster3 = subset(rownames(s.statistics), ct2 == 3)
index3 = match(cluster3, rownames(s.statistics))
clust3mean = colMeans(s.statistics[index3, ])

cluster4 = subset(rownames(s.statistics), ct2 == 4)
index4 = match(cluster4, rownames(s.statistics))
clust4mean = colMeans(s.statistics[index4, , drop = FALSE])

cluster5 = subset(rownames(s.statistics), ct2 == 5)
index5 = match(cluster5, rownames(s.statistics))
clust5mean = colMeans(s.statistics[index5, ])

cluster6 = subset(rownames(s.statistics), ct2 == 6)
index6 = match(cluster6, rownames(s.statistics))
clust6mean = colMeans(s.statistics[index6, ])

cluster7 = subset(rownames(s.statistics), ct2 == 7)
index7 = match(cluster7, rownames(s.statistics))
clust7mean = colMeans(s.statistics[index7, , drop = FALSE])

cluster8 = subset(rownames(s.statistics), ct2 == 8)
index8 = match(cluster8, rownames(s.statistics))
clust8mean = colMeans(s.statistics[index8, , drop = FALSE])

cluster9 = subset(rownames(s.statistics), ct2 == 9)
index9 = match(cluster9, rownames(s.statistics))
clust9mean = colMeans(s.statistics[index9, , drop = FALSE])

cluster10 = subset(rownames(s.statistics), ct2 == 10)
index10 = match(cluster10, rownames(s.statistics))
clust10mean = colMeans(s.statistics[index10, , drop = FALSE])

cluster11 = subset(rownames(s.statistics), ct2 == 11)
index11 = match(cluster11, rownames(s.statistics))
clust11mean = colMeans(s.statistics[index11, , drop = FALSE])

cluster12 = subset(rownames(s.statistics), ct2 == 12)
index12 = match(cluster12, rownames(s.statistics))
clust12mean = colMeans(s.statistics[index12, , drop = FALSE])

clustMeans <- rbind(clust1mean, clust2mean, clust3mean, clust4mean, clust5mean, clust6mean, clust7mean, clust8mean, clust9mean, clust10mean, clust11mean, clust12mean)
clustMeans
library(knitr)
kable(clustMeans, caption = "Column Means for Each Cluster-HC.complete")
```

```{r}
# Interpreting HC.complete based on PCA
plot(pca$scores[, 1:2], col = 'white', main='PCA Biplot - HC.complete')
text(pca$scores[, 1:2], labels=row.names(s.statistics), col=ct2, cex=0.5)
```
This plot still looks confusing.


```{r}
# make a map
country_clusters <- data.frame(country = names(ct), cluster = ct2)
country_clusters$country <- gsub("United States", "USA", country_clusters$country)
country_clusters$country <- gsub("United Kingdom", "UK", country_clusters$country)
country_clusters$country <- gsub("Czech Republic (Czechia)", "Czech Republic", country_clusters$country)
country_clusters$country <- gsub("Côte d'Ivoire", "Ivory Coast", country_clusters$country)
country_clusters$country <- gsub("DR Congo", "Democratic Republic of the Congo", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Trinidad", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Tobago", country_clusters$country)
country_clusters$country <- gsub("State of Palestine", "Palestine", country_clusters$country)
country_clusters$country <- gsub("Congo", "Republic of Congo", country_clusters$country)
country_clusters$country <- gsub("Eswatini", "Swaziland", country_clusters$country)
country_clusters$country <- gsub("Cabo Verde", "Cape Verde", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Antigua", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Barbuda", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Saint Kitts", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Nevis", country_clusters$country)
country_clusters$country <- gsub("St. Vincent & Grenadines", "Saint Vincent", country_clusters$country)
country_clusters$country <- gsub("Sao Tome & Principe", "Sao Tome and Principe", country_clusters$country)

world_map <- map_data("world")

world_map_clusters <- merge(world_map, country_clusters, by.x = "region", by.y = "country", all.x = TRUE)

ggplot(data = world_map_clusters) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(cluster))) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  theme_minimal() +
  labs(fill = "Cluster")
```

## hierarchical clustering - average

```{r}
## hierarchical clustering - average
hc.average <- hclust(dist, 'average')
# cluster dendrogram
plot(hc.average, cex = 0.3, main = 'Cluster Dendrogram of HC.average')
```
```{r}
# find the number of cluster
plot(rev(hc.average$height), main='scree plot of HC.average')

```
Based on the scree plot, choose h = 6 as cut point.
```{r}
ct3 <- cutree(hc.average, h = 6)
table(ct3)
```
```{r}
# cluster mean-HC.average
cluster1 = subset(rownames(s.statistics), ct3 == 1)
index1 = match(cluster1, rownames(s.statistics))
clust1mean = colMeans(s.statistics[index1, , drop = FALSE])

cluster2 = subset(rownames(s.statistics), ct3 == 2)
index2 = match(cluster2, rownames(s.statistics))
clust2mean = colMeans(s.statistics[index2, , drop = FALSE])

cluster3 = subset(rownames(s.statistics), ct3 == 3)
index3 = match(cluster3, rownames(s.statistics))
clust3mean = colMeans(s.statistics[index3, ])

cluster4 = subset(rownames(s.statistics), ct3 == 4)
index4 = match(cluster4, rownames(s.statistics))
clust4mean = colMeans(s.statistics[index4, , drop = FALSE])

cluster5 = subset(rownames(s.statistics), ct3 == 5)
index5 = match(cluster5, rownames(s.statistics))
clust5mean = colMeans(s.statistics[index5, , drop = FALSE])

cluster6 = subset(rownames(s.statistics), ct3 == 6)
index6 = match(cluster6, rownames(s.statistics))
clust6mean = colMeans(s.statistics[index6, ])

cluster7 = subset(rownames(s.statistics), ct3 == 7)
index7 = match(cluster7, rownames(s.statistics))
clust7mean = colMeans(s.statistics[index7, , drop = FALSE])

cluster8 = subset(rownames(s.statistics), ct3 == 8)
index8 = match(cluster8, rownames(s.statistics))
clust8mean = colMeans(s.statistics[index8, , drop = FALSE])

cluster9 = subset(rownames(s.statistics), ct3 == 9)
index9 = match(cluster9, rownames(s.statistics))
clust9mean = colMeans(s.statistics[index9, , drop = FALSE])

cluster10 = subset(rownames(s.statistics), ct3 == 10)
index10 = match(cluster10, rownames(s.statistics))
clust10mean = colMeans(s.statistics[index10, , drop = FALSE])

clustMeans <- rbind(clust1mean, clust2mean, clust3mean, clust4mean, clust5mean, clust6mean, clust7mean, clust8mean, clust9mean, clust10mean)

library(knitr)
kable(clustMeans, caption = "Column Means for Each Cluster-HC.average")
```
```{r}
# Interpreting HC.average based on PCA
plot(pca$scores[, 1:2], col = 'white', main='PCA Biplot - HC.average')
text(pca$scores[, 1:2], labels=row.names(s.statistics), col=ct3, cex=0.5)
```

```{r}
# make a map
country_clusters <- data.frame(country = names(ct), cluster = ct3)
country_clusters$country <- gsub("United States", "USA", country_clusters$country)
country_clusters$country <- gsub("United Kingdom", "UK", country_clusters$country)
country_clusters$country <- gsub("Czech Republic (Czechia)", "Czech Republic", country_clusters$country)
country_clusters$country <- gsub("Côte d'Ivoire", "Ivory Coast", country_clusters$country)
country_clusters$country <- gsub("DR Congo", "Democratic Republic of the Congo", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Trinidad", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Tobago", country_clusters$country)
country_clusters$country <- gsub("State of Palestine", "Palestine", country_clusters$country)
country_clusters$country <- gsub("Congo", "Republic of Congo", country_clusters$country)
country_clusters$country <- gsub("Eswatini", "Swaziland", country_clusters$country)
country_clusters$country <- gsub("Cabo Verde", "Cape Verde", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Antigua", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Barbuda", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Saint Kitts", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Nevis", country_clusters$country)
country_clusters$country <- gsub("St. Vincent & Grenadines", "Saint Vincent", country_clusters$country)
country_clusters$country <- gsub("Sao Tome & Principe", "Sao Tome and Principe", country_clusters$country)

world_map <- map_data("world")

world_map_clusters <- merge(world_map, country_clusters, by.x = "region", by.y = "country", all.x = TRUE)

ggplot(data = world_map_clusters) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(cluster))) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  theme_minimal() +
  labs(fill = "Cluster")
```

## kmeans clustering

```{r}
## kmeans clustering
# find the number of cluster
plot.wgss = function(mydata, maxc = nrow(mydata)-1) {
  wgss = numeric(maxc)
  for (i in 1:maxc){
    km <- kmeans(mydata, centers=i, nstart = 10)
    wgss[i] = km$tot.withinss
  }
  plot(1:maxc, wgss, type="b", xlab="Number of Clusters",
  ylab="Within groups sum of squares", main="Scree Plot") 
}

plot.wgss(s.statistics, 20)
```
Based on the scree plot, k = 3.
```{r}
km <- kmeans(s.statistics, centers = 3,, nstart = 10)
table(km$cluster)
```
```{r}
# cluster mean-kmeans
options(digits = 3)
clusterMeans <- km$centers
rownames(clusterMeans) <- c("Cluster 1", "Cluster 2", "Cluster 3")

library(knitr)
kable(clusterMeans, caption = "Column Means for each Cluster - kmeans")
```
```{r}
pca$loadings[,1:2]
```

```{r}
# Interpreting kmeans based on PCA 
plot(pca$scores[, 1:2], col = 'white', main='PCA Biplot - kmeans')
text(pca$scores[, 1:2], labels=row.names(s.statistics), col=km$cluster, cex=0.4)
```

```{r}
statistics_1 <- as.data.frame(s.statistics)
statistics_1$Cluster <- km$cluster
cluster1_data <- statistics_1[statistics_1$Cluster == 1, ]
cluster1_data
cluster2_data <- statistics_1[statistics_1$Cluster == 2, ]
cluster2_data
cluster3_data <- statistics_1[statistics_1$Cluster == 3, ]
cluster3_data
```


```{r}
# make a map
country_clusters <- data.frame(country = names(km$cluster), cluster = km$cluster)
country_clusters$country <- gsub("United States", "USA", country_clusters$country)
country_clusters$country <- gsub("United Kingdom", "UK", country_clusters$country)
#country_clusters$country <- gsub("Hong Kong", "China", country_clusters$country)
country_clusters$country <- gsub("Czech Republic (Czechia)", "Czech Republic", country_clusters$country)
country_clusters$country <- gsub("Côte d'Ivoire", "Ivory Coast", country_clusters$country)
country_clusters$country <- gsub("DR Congo", "Democratic Republic of the Congo", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Trinidad", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Tobago", country_clusters$country)
#country_clusters$country <- gsub("Macao", "China", country_clusters$country)
country_clusters$country <- gsub("State of Palestine", "Palestine", country_clusters$country)
country_clusters$country <- gsub("Congo", "Republic of Congo", country_clusters$country)
country_clusters$country <- gsub("Eswatini", "Swaziland", country_clusters$country)
country_clusters$country <- gsub("Cabo Verde", "Cape Verde", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Antigua", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Barbuda", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Saint Kitts", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Nevis", country_clusters$country)
country_clusters$country <- gsub("St. Vincent & Grenadines", "Saint Vincent", country_clusters$country)
country_clusters$country <- gsub("Sao Tome & Principe", "Sao Tome and Principe", country_clusters$country)

world_map <- map_data("world")

world_map_clusters <- merge(world_map, country_clusters, by.x = "region", by.y = "country", all.x = TRUE)

ggplot(data = world_map_clusters) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(cluster))) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  theme_minimal() +
  labs(fill = "Cluster")
```

## model-based clustering

```{r}
## model-based clustering
library(mclust)
mc <- Mclust(s.statistics)
summary(mc)
```
Based on the results, there are 3 clusters.


```{r}
# find the number of cluster
plot(mc, what='BIC')
```
```{r}
summary(mc$BIC)
```
Based on the result, we can confirm that the number of clusters is 3.
```{r}
# clusterMeans-ModelBased
clusterMeans <- mc$parameters$mean
clusterMeans <- t(clusterMeans)
rownames(clusterMeans) <- c("Cluster 1", "Cluster 2", "Cluster 3")
library(knitr)
kable(clusterMeans, caption = "Column Means for each Cluster - ModelBased")
```


```{r}
# Interpreting ModelBased Clusters based on PCA
plot(pca$scores[, 1:2], col = 'white', main='PCA Biplot - ModelBased Clustering')
text(pca$scores[, 1:2], labels=row.names(s.statistics), col=mc$classification, cex=0.5)
```
```{r}
statistics_2 <- as.data.frame(s.statistics)
statistics_2$Cluster <- mc$classification
cluster1_data <- statistics_2[statistics_2$Cluster == 1, ]
cluster1_data
cluster2_data <- statistics_2[statistics_2$Cluster == 2, ]
cluster2_data
cluster3_data <- statistics_2[statistics_2$Cluster == 3, ]
cluster3_data
```

```{r}
clust.data = cbind(rownames(mydata), mc$classification,mc$uncertainty)
sorted_uncertainty <- clust.data[order(-mc$uncertainty),]
head(sorted_uncertainty, 10)
```



```{r}
# make a map
country_clusters <- data.frame(country = names(mc$classification), cluster = mc$classification)
country_clusters$country <- gsub("United States", "USA", country_clusters$country)
country_clusters$country <- gsub("United Kingdom", "UK", country_clusters$country)
#country_clusters$country <- gsub("Hong Kong", "China", country_clusters$country)
country_clusters$country <- gsub("Czech Republic (Czechia)", "Czech Republic", country_clusters$country)
country_clusters$country <- gsub("Côte d'Ivoire", "Ivory Coast", country_clusters$country)
country_clusters$country <- gsub("DR Congo", "Democratic Republic of the Congo", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Trinidad", country_clusters$country)
country_clusters$country <- gsub("Trinidad and Tobago", "Tobago", country_clusters$country)
#country_clusters$country <- gsub("Macao", "China", country_clusters$country)
country_clusters$country <- gsub("State of Palestine", "Palestine", country_clusters$country)
country_clusters$country <- gsub("Congo", "Republic of Congo", country_clusters$country)
country_clusters$country <- gsub("Eswatini", "Swaziland", country_clusters$country)
country_clusters$country <- gsub("Cabo Verde", "Cape Verde", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Antigua", country_clusters$country)
country_clusters$country <- gsub("Antigua and Barbuda", "Barbuda", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Saint Kitts", country_clusters$country)
country_clusters$country <- gsub("Saint Kitts & Nevis", "Nevis", country_clusters$country)
country_clusters$country <- gsub("St. Vincent & Grenadines", "Saint Vincent", country_clusters$country)
country_clusters$country <- gsub("Sao Tome & Principe", "Sao Tome and Principe", country_clusters$country)

world_map <- map_data("world")

world_map_clusters <- merge(world_map, country_clusters, by.x = "region", by.y = "country", all.x = TRUE)

ggplot(data = world_map_clusters) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(cluster))) +
  expand_limits(x = world_map$long, y = world_map$lat) +
  theme_minimal() +
  labs(fill = "Cluster")
```


Since we choose different number of clusters for each method, it is not very meaningful to use X-squared to decide which model is better.
When comparing clusterings with different numbers of clusters, the results of the Chi-Squared test should be interpreted with caution.In such cases, it might be more effective to use metrics specifically designed for clustering analysis, like the Silhouette Coefficient or the Calinski-Harabasz Index.

```{r}
## Silhouette Coefficient
library(cluster)
# HC.single
silhouettescores.single <- silhouette(ct, dist(s.statistics))
plot(silhouettescores.single, main = "silhouette scores of HC.single")
# HC.complete
silhouettescores.complete <- silhouette(ct2, dist(s.statistics))
plot(silhouettescores.complete, main = "silhouette scores of HC.complete")
# HC.average
silhouettescores.average <- silhouette(ct3, dist(s.statistics))
plot(silhouettescores.average, main = "silhouette scores of HC.average")
```
```{r}
options(digits = 3)
# kmeans
silhouettescores.kmeans <- silhouette(km$cluster, dist(s.statistics))
summary(silhouettescores.kmeans)
plot(silhouettescores.kmeans, main = "silhouette scores of kmeans clustering")
```
```{r}
# Model Based
silhouettescores.mc <- silhouette(mc$classification, dist(s.statistics))
summary(silhouettescores.mc)
plot(silhouettescores.mc, main = "silhouette scores of Model Based clustering")
```


Values range from -1 to 1. A high value indicates that the object is well matched to its own cluster and poorly matched to neighboring clusters.



